Cover Your Basis: Comprehensive Data-Driven Characterization of the Binary Black Hole Population

### 2022-10-11 src/tex/methods.tex
update some methods and results

We construct our data-driven model with the application of basis splines, or B-Splines \citep{deBoor78}. B-Splines of order $k$, are a set of order $k-1$ polynomials that 
span the space of possible spline functions interpolated from a given set of knot locations. The basis representation of the splines allows for the computational 
expensive interpolation to be done in a single preprocessing step -- amortizing the cost of model evaluation during inference. To mitigate the unwanted side effects of 
adding extra knots and to avoid running a model grid of differing numbers of knots (as in \citet{Edelman_2022ApJ}), we use the smoothing prior for Bayesian P-Splines, 
allowing the data to pick the optimal scale needed to fit the present features \citep{eilers2021practical,BayesianPSplines,Jullion2007RobustSO}. We discuss basis splines, the smoothing prior 
and our specific prior choices on hyperparameters in Appendix \ref{sec:basis_splines}, \ref{sec:psplines} and \ref{sec:modelpriors}.

We parameterize each binaries' masses with the primary (more massive component) mass 
and the mass ratio ($q=\frac{m_2}{m_1}$), bounded in the range from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger with each component spin magnitude $a_1$ and $a_2$, along with the cosine of the tilt angle of each component, 
$\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binaries orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. For the primary mass distribution, we model the log probability with a B-Spline interpolated over 
knots linearly spaced in $\log(m_1)$ in the range of support given our observations ($[6.5\msun-100\msun]$), such that $\log(p(m_1 | \bm{c})) \propto B_{k}(\log(m_1) | \bm{c})$, where $B_{k}$ is the 
$k$-th order B-Spline function with basis coefficients, $\bm{c}$. We follow the same procedure for the models in mass ratio and spin distributions with knots spaced linearly across each domain, 
so that we have $\log(p(\theta | \bm{c}_\theta)) \propto B_{k}(\theta | \bm{c}_\theta)$, where $\theta$ can be $q$, $a_1$, $a_2$, $\cos{\theta_1}$ or $\cos{\theta_2}$. 
For the spin magnitude and tilt distributions we construct two versions of the model: first, we model  
each component's distribution as independently and identically distribution (IID), where we have a single B-Spline model and parameters (coefficients) for each binary spin. 
Secondly, we model each component's distribution to be unique, fitting separate sets of coefficients for the B-Spline models of the primary and secondary spin distributions. 
Lastly, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and it's redshift, which we now use interchangeably. 
We take a semi-parametric approach to model the evolution of the merger rate with redshift, following \citet{Edelman_2022ApJ}, that parameterizes modulations to an underlying model 
with splines (in our case basis Splines). We use the \textsc{PowerlawRedshift} model as the underlying distribution to modulate, which has a single hyperparameter, $\lambda_z$, and 
probability density defined as: $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. For more detailed descriptions of each model and 
specific prior choices used for the hyperparameters fit in Appendix \ref{sec:modelpriors}. Now that we have our comprehensive data-driven population model over 
each of the 7 modeled parameters built, we fit our model using the usual hierarchical Bayesian inference framework (see appendix \ref{sec:hierarchical_inference} for a review) in order to characterize 
the population of BBHs in the most recent catalog of GW observations, GWTC-3 \citep{GWTC3,o3b_astro_dist}.

>>>>

We construct our data-driven model with the application of basis splines, or B-Splines \citep{deBoor78}. B-Splines of order $k$ are a set of order $k-1$ polynomials that 
span the space of possible spline functions interpolated from a given set of knot locations. The basis representation of the splines allows for the computationally
expensive interpolation to be done in a single preprocessing step -- amortizing the cost of model evaluation during inference. To mitigate the unwanted side effects of 
adding extra knots and to avoid running a model grid of differing numbers of knots (as in \citet{Edelman_2022ApJ}), we use the smoothing prior for Bayesian P-Splines \citep{eilers2021practical,BayesianPSplines,Jullion2007RobustSO}, 
allowing the data to pick the optimal scale needed to fit the present features. We discuss basis splines, the smoothing prior, and our specific prior choices on hyperparameters in Appendix \ref{sec:basis_splines}, \ref{sec:psplines} and \ref{sec:modelpriors}.

We parameterize each binaries' masses with the primary (more massive component) mass ($m_1$) 
and the mass ratio ($q=m_2/m_1$) with support from 0 to 1. Furthermore, we model 4 of the 6 total 
spin degrees of freedom of a binary merger: component spin magnitudes $a_1$ and $a_2$, and (cosine of) the tilt angles of each component, $\cos{\theta_1}$ and $\cos{\theta_2}$. The tilt angle is defined as the angle between each components' spin vector and the binary's orbital angular momentum vector. 
We assume the polar spin angles are uniformly distributed in the orbital plane. For the primary mass distribution, we model the log probability with a B-Spline interpolated over 
knots linearly spaced in $\log(m_1)$ in the range of support given our observations ($[6.5\msun-100\msun]$), such that $\log(p(m_1 | \bm{c})) \propto B_{k}(\log(m_1) | \bm{c})$, where $B_{k}$ is the 
$k$-th order B-Spline function with basis coefficients $\bm{c}$. We follow the same procedure for the models in mass ratio and spin distributions with knots spaced linearly across each domain 
so that we have $\log(p(\theta | \bm{c}_\theta)) \propto B_{k}(\theta | \bm{c}_\theta)$, where $\theta$ can be $q$, $a_1$, $a_2$, $\cos{\theta_1}$ or $\cos{\theta_2}$. 
For the spin magnitude and tilt distributions we construct two versions of the model: first, we model  
each component's distribution as independently and identically distribution (IID), where we have a single B-Spline model and parameters (coefficients) for each binary spin. 
Secondly, we model each component's distribution to be unique, fitting separate sets of coefficients for the B-Spline models of the primary and secondary spin distributions. 
Lastly, we fit a population model on the redshift or luminosity distance distribution of BBHs, assuming a $\Lambda\mathrm{CDM}$ cosmology defined by the parameters 
from the Planck 2015 results \citep{Planck2015}. This defines an analytical mapping between each event's inferred luminosity distance, and its redshift, which we now use interchangeably. 
We take a semi-parametric approach to model the evolution of the merger rate with redshift, following \citet{Edelman_2022ApJ}, that parameterizes modulations to an underlying model 
with splines (in our case basis splines). We use the \textsc{PowerlawRedshift} model as the underlying distribution to modulate, which has a single hyperparameter, $\lambda_z$, and 
probability density defined as: $p(z|\lambda_z)\propto \frac{dV_c}{dz}(1+z)^{\lambda_z-1}$ \citep{Fishbach_2018redshift}. For more detailed descriptions of each model and 
specific prior choices used for the hyperparameters see Appendix \ref{sec:modelpriors}. Now that we have our comprehensive data-driven population model over 
each of the 7 modeled parameters built, we fit our model using the usual hierarchical Bayesian inference framework (see appendix \ref{sec:hierarchical_inference} for a review) in order to characterize 
the population of BBHs in the most recent catalog of GW observations, GWTC-3 \citep{GWTC3,o3b_astro_dist}.

### 2022-10-11 src/tex/results.tex
fix dash range
update some methods and results

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneously infer the astrophysical mass, spin and redshift distributions of 
binary black holes (BBHs) given a catalog of gravitational wave observations. Following the same cut on the recent GWTC-3 catalog done in the LVK's 
accompanying BBH population analysis, we have 70 possible BBH mergers with false alarm rates less than 1 per year \citep{GWTC3,o3b_astro_dist}. Since it was concluded to be an outlier of the rest of 
the BBH population in both GWTC-2 and GWTC-3, we choose to omit the poorly understood event, GW190814 \citep{190814disc,o3a_pop,o3b_astro_dist,Essick_2022}. This leaves us with 
a catalog of 69 confident BBH mergers, observed over a period of about 2 years, from which we want to infer its population properties. Our study provides the first comprehensive 
data-driven investigation of the BBH mass, spin and redshift distributions, uncovering new insights and corroborating those found with other methods. 
We start by looking towards our model's inferences of the mass distribution.

\begin{figure*}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/mass_distribution_plot.pdf}
        \caption{The marginal primary mass distribution inferred with the B-Spline model (red), with 64 knots spaced linearly in $\log m_1$, from 6.5\msun 
        to 100\msun. The solid line shows the population predictive distribution (PPD) and the shaded region, 
        the 90\% credible interval. We show the inferred PPD from the \textsc{PowerlawPeak} (blue) and \textsc{PowerlawSpline} (green) models 
        from the LVK's GWTC-3 population analyses \citep{o3b_astro_dist}.}
        \label{fig:mass_distribution}
    \end{centering}
    \script{mass_distribution_plot.py}
\end{figure*}

@@

Figure \ref{fig:mass_distribution} shows the primary mass distribution inferred with our B-Spline model (red), where we see that it recovered consistent 
features with what was inferred with the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \citep{Talbot_2018,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our B-Spline model finds a peak in merger rate density as a function of primary mass, at both $\sim10\msun$ and $\sim35\msun$ agreeing with those 
reported using the same dataset in \citet{o3b_astro_dist}. The B-Spline model finds the same feature at $\sim18\msun$ as the \textsc{PowerlawSpline} model, but also 
agreeing with the \textsc{PowerlawSpline} model, the mass distribution is more uncertain and less constrained in this region than other two confident peaks found.
Additionally, there is some disagreement at low masses where the models based on power laws show a much higher rate below $\sim8--9\msun$. This is partly due to  
the minimum mass hyperparameter, which is where the power law ``begins'' in those models. They enforce this hyperparameter as the minimum allowable primary and secondary 
masses of the catalog. This leads to inferences of $m_\mathrm{min}$ below the minimum observed primary mass in the catalog, $\sim 6.4\msun$, to account for secondary BBH masses 
that can possibly get to as low as $\sim 3\msun$. Since non-parametric models do not provide meaningful insights where there is no data - we enforce the minimum primary mass of 
our B-Spline model to be $6.5\msun$, while secondary BBH masses are enforced to be greater than $3\msun$. The lack of observations in the low primary mass region could 
point towards signs of a decrease in merger rate from a purported ``lower mass gap'' between neutron star and BH masses, or could be a feature of selection effects which make it 
more unlikely to observe BBHs in that region of parameter space due to small sensitive volumes probed. We also find no evidence for a sharp fall off in merger rate either following 
the pileup at $\sim35\,\msun$, which is expected if such a pileup was due to pulsational pair instability supernovae (PPISNe), or where the maximum mass truncation of the power law models 
is inferred. The lack of any high mass truncation along with the most significant peak found to be at $\sim35\msun$, significantly lower than expected, may pose challenges for conventional stellar 
evolution theory. This could be hinting at the significant presence of subpopulations that avoid PISN during binary formation, but the confirmation of the existence of such subpopulations 
cannot be determined with current catalog sizes.

>>>>

We use hierarchical Bayesian inference (see Appendix \ref{sec:hierarchical_inference}) to simultaneously infer the astrophysical mass, spin, and redshift distributions of 
binary black holes (BBHs) given a catalog of gravitational wave observations. Following the same cut on the recent GWTC-3 catalog done in the LVK's 
accompanying BBH population analysis, we have 70 possible BBH mergers with false alarm rates less than 1 per year \citep{GWTC3,o3b_astro_dist}. Since it was concluded to be an outlier of the rest of 
the BBH population in both GWTC-2 and GWTC-3, we choose to omit the poorly understood event, GW190814 \citep{190814disc,o3a_pop,o3b_astro_dist,Essick_2022}. This leaves us with 
a catalog of 69 confident BBH mergers, observed over a period of about 2 years, from which we want to infer population properties. Our study provides the first comprehensive 
data-driven investigation of the BBH mass, spin, and redshift distributions, uncovering new insights and corroborating those found with other methods. 
We start with our inference of the mass distribution.

\begin{figure*}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/mass_distribution_plot.pdf}
        \caption{The marginal primary mass distribution inferred with the B-Spline model (red), with 64 knots spaced linearly in $\log m_1$, from 6.5\msun 
        to 100\msun. The solid line shows the population predictive distribution (PPD), and the shaded region 
        the 90\% credible interval. We show the inferred PPD from the \textsc{PowerlawPeak} (blue) and \textsc{PowerlawSpline} (green) models 
        from the LVK's GWTC-3 population analyses \citep{o3b_astro_dist}.}
        \label{fig:mass_distribution}
    \end{centering}
    \script{mass_distribution_plot.py}
\end{figure*}

@@

Figure \ref{fig:mass_distribution} shows the primary mass distribution inferred with our B-Spline model (red), where we see features consistant with those inferred by the \textsc{PowerlawPeak} and \textsc{PowerlawSpline} mass models \citep{Talbot_2018,o3a_pop,Edelman_2022ApJ,o3b_astro_dist}. 
In particular our B-Spline model finds peaks in merger rate density as a function of primary mass at both $\sim10\msun$ and $\sim35\msun$, agreeing with those 
reported using the same dataset in \citet{o3b_astro_dist}. The B-Spline model finds the same feature at $\sim18\msun$ as the \textsc{PowerlawSpline} model, but remains consistent with the \textsc{PowerlawSpline} model; the mass distribution is more uncertain in this region.
We find the largest disagreement at low masses, where the power-law-based models show a higher rate below ${\sim}8-9\msun$. This is partly due to the minimum mass hyperparameter (where the power law ``begins'') serving as the minimum allowable primary and secondary masses of the catalog. This leads to inferences of $m_\mathrm{min}$ below the minimum observed primary mass in the catalog, $\sim 6.4\msun$, to account for secondary BBH masses 
that can possibly get as low as $\sim 3\msun$. We choose a minimum primary mass for 
the B-Spline model to be $6.5\msun$, while secondary masses are enforced to be greater than $3\msun$. The lack of observations of binaries with low primary mass make rate estimates in this region strongly model and prior dependent.  We could be seeing signs of a decrease in merger rate from a ``lower mass gap'' between neutron star and BH masses, or we could be seeing fluctuations due to low-number statistics.  Either way we explect this to be resolved with future catalog updates. We also find no evidence for a sharp fall off in merger rate either following 
the pileup at $\sim35\,\msun$ -- expected if such a pileup was due to pulsational pair instability supernovae (PPISNe) -- or where the maximum mass truncation of the power law models 
are inferred. The lack of any high mass truncation, along with the peak at $\sim35\msun$ (significantly lower than expected from PPISNe) may pose challenges for conventional stellar 
evolution theory. This could be hinting at the presence of subpopulations that avoid pair instability supernovae during binary formation, but the confirmation of the existence of such subpopulations 
cannot be determined with the current catalog.
